---
title: "p8105_hw5_at3583"
author: "Alexandria D.Taylor,"
date: "`r format(Sys.Date())`"
output: github_document
---

**This homework applied Monte Carlo methods, hypothesis testing, and proportion confidence intervals to examine probability, power, and uncertainty in simulated and real-world data.**

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE, 
    warning = FALSE
)
library(tidyverse)
library(broom)
library(purrr)

# Global color palette
primary_color  <- "springgreen3"
accent_color   <- "springgreen4"

# Global ggplot theme
theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "bottom"
    )
)
```

# Problem 1 -- Birthday Simulation

```{r}
# Generate birthday simulation function

birthday_sim = function(n) {
  birthdays = sample(1:365, size = n, replace = TRUE)
  any(duplicated(birthdays))
}

# Run birthday simulation function for group sizes 2 to 50 at 10,000 times each

set.seed(123)

group_sizes = 2:50

simulate_results = tibble(
  group_size = group_sizes,
  prob_shared_birthday = map_dbl(group_sizes, function(n) {
    mean(replicate(10000, birthday_sim(n)))
    
  })
)

# Data summary:compute the smallest group size where probability >= 0.5
threshold_50 <- simulate_results %>%
  filter(prob_shared_birthday >= 0.5) %>%
  slice(1) %>%
  pull(group_size)
```

```{r}

# Probability vs group size plot

simulate_results %>%
  ggplot(aes(x = group_size, y = prob_shared_birthday)) +
  geom_line(color = primary_color, size = 1.2) +
  geom_point(color = accent_color, size = 2) +
  labs(
    title = "Probability of Shared Birthday vs Group Size",
    x = "Group Size (n)",
    y = "Probability of at least one shared birthday"
  ) +
  theme_minimal()
```

Summary: The likelihood that two people in a group share the same birthday increases rapidly with group size. Despite there being 365 possible birth dates, the probability exceeds 50% once the group size reaches **`r threshold_50` people**, a counterintuitive result known as the *birthday paradox*. By around **50 individuals**, the chance is nearly certain, making a shared birthday almost guaranteed.

# Problem 2 -- One Sample t-Test

```{r}
# Generate t-test parameters

set.seed(8105)

n_sims <- 5000
n      <- 30
sigma  <- 5

# True means test 
mus    <- 0:6 
alpha  <- 0.05
```

```{r}
# Set-up n_sims dataset simultation at true mean mu

simulate_for_mu <- function(mu, n, sigma, n_sims, alpha = 0.05) {
  purrr::rerun(n_sims, rnorm(n, mean = mu, sd = sigma)) |>
    purrr::set_names(NULL) |>
    purrr::map_dfr(function(x) {
      tt <- t.test(x, mu = 0)
      tibble(
        mu      = mu,
        mu_hat  = mean(x),
        p_val   = broom::tidy(tt)$p.value,
        reject  = p_val < alpha
      )
    })
}

# Run n_sims dataset simulatation 

sim_results <- purrr::map_dfr(mus, simulate_for_mu, n = n, sigma = sigma, n_sims = n_sims, alpha = alpha)
```

## <small><em> Power by effect size (mu) </small></em>

```{r}
# Generate power data frame from simulation results
power_df <- sim_results |>
  group_by(mu) |>
  summarise(power = mean(reject), .groups = "drop")

# Generate average estimates -- overall vs. conditional on reject
estimates_df <- sim_results |>
  group_by(mu) |>
  summarise(
    mean_hat_all    = mean(mu_hat),
    mean_hat_reject = mean(mu_hat[reject]),
    .groups = "drop"
  ) |>
  tidyr::pivot_longer(
    cols = c(mean_hat_all, mean_hat_reject),
    names_to = "subset",
    values_to = "mean_hat"
  ) |>
  mutate(subset = recode(subset,
                         mean_hat_all = "All samples",
                         mean_hat_reject = "Only rejected H0"))
```

```{r}
# Plot power df
power_df |>
  ggplot(aes(x = mu, y = power)) +
  geom_line(color = primary_color, linewidth = 1.3) +
  geom_point(color = accent_color, size = 2.2) +
  scale_x_continuous(breaks = mus) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    title = "Power of one-sample t-test (n = 30, σ = 5, α = 0.05)",
    x = "True mean (μ)",
    y = "Power (Pr reject H₀: μ = 0)"
  ) +
  theme_minimal()
```

Summary: The probability of rejecting the null hypothesis increases as the true mean (μ) increases. When μ is close to **0**, the test seldom rejects the null, indicating low power. As μ becomes larger, the effect becomes more detectable relative to noise, and power rises rapidly toward **1**. This illustrates the fundamental principle that the expected relationship that larger effect sizes yield greater power in hypothesis testing.

```{r}
# Plot-estimates dataframe
estimates_df |>
  ggplot(aes(x = mu, y = mean_hat, color = subset)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = mus) +
  scale_color_manual(values = c("All samples" = primary_color,
                               "Reject H₀ only" = accent_color)) +
  labs(
    title = "Average estimate of μ̂ vs true μ",
    subtitle = "Conditioning on rejection inflates μ̂ (selection bias)",
    x = "True mean (μ)",
    y = "Average μ̂",
    color = ""
  ) +
  theme_minimal()
```

Summary: Across all simulated samples, the average estimate of μ̂ closely matches the true mean μ, indicating that the estimator is unbiased. However, when considering only samples where the null hypothesis was rejected, the average μ̂ is inflated above the true value. This occurs because under low power, only the more extreme sample means yield significant results, introducing selection bias.

# Problem 3 -- Washington Post

```{r}
library(tidyverse)
library(janitor)
library(broom)
library(purrr)

# Load and clean homicide data
homicide_df <- read_csv("https://raw.githubusercontent.com/washingtonpost/data-homicides/master/homicide-data.csv") |>
  clean_names()

# Create city_state variable & flag unsolved cases
homicide_city <- homicide_df |> 
  mutate(
    city_state = str_c(city, ", ", state),
    unsolved = disposition %in% c("Closed without arrest", "Open/No arrest")
  )
city_summary <- homicide_city |> 
  group_by(city_state) |> 
  summarise(
    total_homicides = n(),
    unsolved_homicides = sum(unsolved),
    .groups = "drop"
  )

knitr::kable(city_summary, caption = "Homicide counts and unsolved cases by city")

```

Summary: The dataset includes **`r nrow(homicide_df)`** homicide records from 50 U.S. cities, with **`r ncol(homicide_df)`** variables detailing victim demographics (age, sex, race), geographic information, and case outcomes. Each row represents a single homicide incident. Some fields contain missing data -- **`r sum(is.na(homicide_df$victim_race))`** cases lack recorded victim race—and several disposition categories required recoding to determine whether a case is solved. In this analysis, both *Open/No arrest* and *Closed without arrest* are classified as unsolved.

## Baltimore, MD

```{r}
# Filter data to Baltimore
baltimore_df <- homicide_city |> 
  filter(city_state == "Baltimore, MD")

# Run prop test
baltimore_test <- prop.test(
  x = sum(baltimore_df$unsolved),
  n = nrow(baltimore_df)
)

# Tidy filtered results
baltimore_summary <- baltimore_test |> 
  broom::tidy() |> 
  select(estimate, conf.low, conf.high)

baltimore_summary
```

The estimated proportion of unsolved homicides in Baltimore is **`r round(baltimore_summary$estimate, 3)`**, with a 95% confidence interval of **(`r round(baltimore_summary$conf.low, 3)`, `r round(baltimore_summary$conf.high, 3)`)**.

## City Analysis, prop.test()

```{r}
# Prop.test for each city
city_prop_tests <- homicide_city %>%
  group_by(city_state) %>%
  summarise(
    total_homicides = n(),
    unsolved_homicides = sum(unsolved),
    .groups = "drop"
  ) %>%
  mutate(
    test_result = map2(unsolved_homicides, total_homicides,
                       ~ prop.test(.x, .y))
  ) %>%
  mutate(
    tidy_result = map(test_result, broom::tidy)
  ) %>%
  select(city_state, tidy_result) %>%
  unnest(tidy_result) %>%
  select(city_state, estimate, conf.low, conf.high)

head(city_prop_tests)
```

```{r}
# Plot city prop test output
city_prop_tests %>%
  ggplot(aes(x = reorder(city_state, estimate), y = estimate)) +
  geom_point(color = accent_color, size = 3) +
  geom_errorbar(aes(ymin = conf.low, ymax = conf.high),
                width = 0.2, color = primary_color) +
  coord_flip() +
  labs(
    title = "Estimated Proportion of Unsolved Homicides by City",
    x = "City",
    y = "Proportion Unsolved (95% CI)"
  ) +
  theme_minimal(base_size = 12) +
  theme(
  axis.text.x = element_text( size = 12,
    margin = margin(r = 30)
   )
  )
```

Summary: Across the 50 cities, the proportion of homicides that remain unsolved varies substantially, ranging from **`r round(min(city_prop_tests$estimate), 2)`** to **`r round(max(city_prop_tests$estimate), 2)`**. Cities with higher unsolved proportions tend to have wider confidence intervals, reflecting greater uncertainty that often stems from smaller total case counts. Baltimore falls among the cities with a notably high share unsolved cases.
