---
title: "p8105_hw3_at3583"
author: "Alexandria D.Taylor,"
date: "`r format(Sys.Date())`"
output: github_document
---

**<small><em>Overview: This homework applied .</small></em>**

```{r setup, include=FALSE}
knitr::opts_chunk$set(
    echo = TRUE,
    message = FALSE, 
    warning = FALSE
)
library(tidyverse)
library(broom)
library(purrr)

# Global color palette
primary_color  <- "springgreen3"
accent_color   <- "springgreen4"

# Global ggplot theme
theme_set(
  theme_minimal(base_size = 12) +
    theme(
      plot.title = element_text(face = "bold"),
      legend.position = "bottom"
    )
)
```

# Problem 1 -- Birthday Simulation

```{r}
#generate & run birthday simulation function

birthday_sim = function(n) {
  birthdays = sample(1:365, size = n, replace = TRUE)
  any(duplicated(birthdays))
}

#run birthday simulation function for group sizes 2 to 50 at 10,000 times each

set.seed(123)

group_sizes = 2:50

simulate_results = tibble(
  group_size = group_sizes,
  prob_shared_birthday = map_dbl(group_sizes, function(n) {
    mean(replicate(10000, birthday_sim(n)))
    
  })
)

#data summary:compute the smallest group size where probability >= 0.5
threshold_50 <- simulate_results %>%
  filter(prob_shared_birthday >= 0.5) %>%
  slice(1) %>%
  pull(group_size)
```

```{r}

#probability vs group size plot

simulate_results %>%
  ggplot(aes(x = group_size, y = prob_shared_birthday)) +
  geom_line(color = primary_color, size = 1.2) +
  geom_point(color = accent_color, size = 2) +
  labs(
    title = "Probability of Shared Birthday vs Group Size",
    x = "Group Size (n)",
    y = "Probability of at least one shared birthday"
  ) +
  theme_minimal()
```

Summary: The likelihood that two people in a group share the same birthday increases rapidly with group size. Despite there being 365 possible birth dates, the probability exceeds 50% once the group size reaches 
**`r threshold_50` people**, a counterintuitive result known as the *birthday paradox*. 
By around **50 individuals**, the chance is nearly certain, making a shared birthday almost guaranteed.

# Problem 2 --  One Sample t-test

```{r}
#generate t-test parameters

set.seed(8105)

n_sims <- 5000
n      <- 30
sigma  <- 5

#true means to test 
mus    <- 0:6 
alpha  <- 0.05
```

```{r}
# set-up n_sims dataset simultation at true mean mu

simulate_for_mu <- function(mu, n, sigma, n_sims, alpha = 0.05) {
  purrr::rerun(n_sims, rnorm(n, mean = mu, sd = sigma)) |>
    purrr::set_names(NULL) |>
    purrr::map_dfr(function(x) {
      tt <- t.test(x, mu = 0)
      tibble(
        mu      = mu,
        mu_hat  = mean(x),
        p_val   = broom::tidy(tt)$p.value,
        reject  = p_val < alpha
      )
    })
}

# run n_sims dataset simultation 

sim_results <- purrr::map_dfr(mus, simulate_for_mu, n = n, sigma = sigma, n_sims = n_sims, alpha = alpha)
```

## Power by effect size (mu)
```{r}
#generate power datafram from simulation results
power_df <- sim_results |>
  group_by(mu) |>
  summarise(power = mean(reject), .groups = "drop")

#generate average estimates-- overall vs. conditional on reject
estimates_df <- sim_results |>
  group_by(mu) |>
  summarise(
    mean_hat_all    = mean(mu_hat),
    mean_hat_reject = mean(mu_hat[reject]),
    .groups = "drop"
  ) |>
  tidyr::pivot_longer(
    cols = c(mean_hat_all, mean_hat_reject),
    names_to = "subset",
    values_to = "mean_hat"
  ) |>
  mutate(subset = recode(subset,
                         mean_hat_all = "All samples",
                         mean_hat_reject = "Only rejected H0"))
```

```{r}
#plot power df
power_df |>
  ggplot(aes(x = mu, y = power)) +
  geom_line(color = primary_color, linewidth = 1.3) +
  geom_point(color = accent_color, size = 2.2) +
  scale_x_continuous(breaks = mus) +
  coord_cartesian(ylim = c(0, 1)) +
  labs(
    title = "Power of one-sample t-test (n = 30, σ = 5, α = 0.05)",
    x = "True mean (μ)",
    y = "Power (Pr reject H₀: μ = 0)"
  ) +
  theme_minimal()
```
Summary: The probability of rejecting the null hypothesis increases as the true mean (μ) increases. 
When μ is close to 0, the test seldom rejects the null, indicating low power. As μ becomes larger, the effect becomes more detectable relative to noise, and power rises rapidly toward 1. This illustrates the fundamental principle that  the expected relationship that larger effect sizes yield greater power in hypothesis testing.
```{r}
# plot-estimates dataframe
estimates_df |>
  ggplot(aes(x = mu, y = mean_hat, color = subset)) +
  geom_line(linewidth = 1) +
  geom_point(size = 2) +
  scale_x_continuous(breaks = mus) +
  scale_color_manual(values = c("All samples" = primary_color,
                               "Reject H₀ only" = accent_color))
  labs(
    title = "Average estimate of μ̂ vs true μ",
    subtitle = "Conditioning on rejection inflates μ̂ (selection bias)",
    x = "True mean (μ)",
    y = "Average μ̂",
    color = ""
  ) +
  theme_minimal()
```
Summary:Across all simulated samples, the average estimate of μ̂ closely matches the true mean μ, indicating that the estimator is unbiased. However, when considering only samples where the null hypothesis was rejected, the average μ̂ is inflated above the 
true value. This occurs because under low power, only the more extreme sample means yield produce significant results, introducing selection bias.